\section*{Acknowledgements}
This thesis is the result of my work at the Department of Ecoinformatics, Biometrics and Forest Growth of the Büsgen Institute of the University of Göttingen between September 2009 and November 2012. It would not have been possible without the support of the following persons, whom I thank herewith.\\
\textbf{tbd:} grote et al ist 2x im LiteraturVZ; Block et al ebenfalls\\
Nach der Überschrift im Header ist ein Punkt. Bei NVL ist das auch, bei JS nicht.


\renewcommand{\labelitemi}{--}
\begin{itemize}
	\item Prof. Dr. Jürgen Nagel acts as referee.
	\item Prof. Dr. Bernhard Möhring acts as co-referee.
	\item Michael Henke, Robert S. Nuske, Tim Ritter, Dr. Rainer Schulz and Jan C. Thiele enriched my work through fruitful discussions, proofreading and technical support.
	\item Jan Hansen helped me when problems with the program WaldPlaner 2.0 occured.
	\item Marco Bender and Jonas Ide helped me through the development of an optimisation algorithm for clustering of sampling points.
	\item The (former) colleagues at the Department of Ecoinformatics, Biometrics and Forest Growth, PD Dr. Wilfried Hakes, Reinhard Hemmerling, Prof. Dr. Winfried Kurth, Dr. Irina Kuzyakova, Dr. Reinhold Meyer, Yongzhi Ong, Sebastian Schoneberg, Prof. Dr. Dr. h.c. Branislav Sloboda, Dr. Katar\'{i}na Smole\v{n}ov\'{a} and Ilona Watteler-Spang always helped me and provided a nice atmosphere, which made me enjoying the work.
	\item The colleagues from the departments of Ecosystem Modelling, Forest Economics and Forest Utilization, Silviculture and Forest Ecology of the Temperate Zones, and Forest Inventory and Remote Sensing made my coming to work gladly.
  \item My family and friends.
\end{itemize}