\chapter{Conclusions}
\label{chap:discussion}
\newpage
\noindent
The aim of this thesis is the evaluation, development and application of robust methods to strengthen the raw material supply of a growing bio-economy sector. It is shown that the success of bio-based companies depends, in parts, on decisions that are made by foresters. Supporting decisions in forest management purposes can, therefore, be beneficial, not only for the forest sector itself but also for wood processing companies. In this thesis, distinct methods to support decisions in forest planning are presented. All enable the reasonable and effective stand treatment planning in order to assess the entire wood potentials of forest stands. This is, of course, also relevant for companies which depend on wood as raw materials. In a wood scarcity scenario, additional wood potentials can have substantial consequences for the success of wood processing companies. The same appears for decisions on the wood distribution. It is not the task of the forest sector to serve all possible consumers with raw materials but since decisions about the wood distribution can cause serious consequences for companies of an economically important sector (Sections \ref{sec:intro:biecon} and \ref{sec:hzb:Konsequenzen}), decisions should be taken very thoroughly.

This reinforces the importance of DSS in the forest management decision process, as they can be used to structure the planning process of wood harvesting and wood distribution into solvable sub-problems \citep[p. 1065-1067]{pretzsch_2008}. The benefits, disadvantages and results of the distinct statistical models for decision support are discussed in the following.

\section{Findings of the thesis}
\label{sec:discussion:findings}
The introductory hypothesis, that matching demands of the rising bio-economy is actually problematic, was verified in Chapter \ref{chap:hzb}. Scarcity of woody biomass and the high complexity of the entire wood supply chain are problems that forest enterprises and bio-economy companies have to cope with. A descriptive analysis of recent wood potentials and wood usages reveals the significance of the research question and thus confirms the need for profound systems to help solve the anticipated raw wood distribution problem. After confirmation of the relevance of the introductory question in Chapter \ref{chap:hzb}, the normative analysis in Chapters \ref{chap:bm} to \ref{chap:opt} examine particular reasons why decision-makers may not be able to exploit the wood potential fully and how optimal potentials could be calculated.

\subsection{Analyzing status and development of raw wood availability in European beech-dominated central Germany}
\label{subsec:discussion:struct:hzb}
The available wood potential of all European beech wood assortments was almost completely exhausted in central Germany in the period between 2002 and 2012 (Chapter \ref{chap:hzb}). Though the European beech wood potential is expected to rise due to ongoing forest development programs, competition on the wood market will probably rise. The only way for upcoming bio-economy companies to establish on the wood market is, thus, to compete with existing market participants. A detailed analysis of the wood market is, therefore, a prerequisite for the success of wood processing companies.

The added value of stem wood is higher than those of any other assortment \citep{nagel_2008}. The stem wood supply is currently entirely exhausted by long-established wood processing companies. It is therefore very difficult for novel companies to establish immediately on the stem wood market. Smaller-dimensioned, low-valued wood assortments make up about 60 \% of the available European Beech wood (Subsection \ref{subsec:hzb:Ergebnisse:Nachhaltig}). Although this low-valued wood is currently almost entirely used as well, the potential seems to be reachable in the future. Approximately half of the wood potential of smaller dimension wood is directly used energetically. If a bio-economy company is able to compete with industrial wood and firewood prices, a solid base of renewable biomass from forests will then become available. This would be beneficial for the forest sector as well because the added value of sorting would increase \cite[p. 67]{mohring_1997}.

Two lessons should be learned from the predicted wood scarcity. First of all, the achievable sustainable wood potential should be estimated as accurately as possible in order to evaluate sources properly. Accurate estimations may uncover currently unused resources. Secondly, decisions on the distribution of the available wood potential must be made very carefully because the success of wood processing companies crucially depends on continuous wood supply quantities. Decisions on distribution of sparse resources can hence have serious consequences for forestry enterprises and for wood processing companies.

\subsection{Biomass functions and nutrient contents of European beech, oak, sycamore maple and ash and their meaning for the biomass supply chain}
\label{subsec:discussion:struct:bm}
The biomass and nutrient contents models from Chapter \ref{chap:bm} can help estimating the ecologically viable wood potential of mixed deciduous forest stands. A comparison of several existing models revealed that the introduced models can significantly improve the estimation accuracy of biomass and nutrient contents in mixed deciduous tree species stands. This might help in identifying, and harvesting, formerly unused potential. The essay is particular relevant for the bio-based sector as innovative production methods often require biomass from deciduous tree species \citep[p. 1]{auer_2016} and the frequency of mixed deciduous stands is increasing steadily \citep{ti_2014}. The models enhance the accuracy of tree-specific biomass estimations. They can thus strengthen the planning accuracy of entire biomass supply chains.

\subsection{Modelling the economically viable wood in the crown of European beech trees}
\label{subsec:discussion:struct:beech_crowns}
The models from Chapter \ref{chap:beech_crowns} are aimed at assessing the optimal wood volume in the crowns of European beech. They can thus be used to predict the full economically viable potential of smaller wood assortments. Allometric analysis revealed large wood potential in the crowns of deciduous trees. It is therefore essential to consider the predicted wood volume from the crown, a by-product of stem wood production. As the dimension of the wood assortments plays a minor role for many activities (Subsection \ref{subsec:intro:struct:bm}), gathering further wood potential from deciduous crowns would seem to be of interest for bio-economy companies. The models also allow a precise prediction of harvestable wood volume prior to harvesting. They can therefore enhance the forecasting of volume flows in the biomass supply chain.

The model is able to calculate the maximal wood potential from an economic perspective only. It is therefore a purely normative-based model that predicts a rational, rather than a realistic, wood potential. Further economic or non-economic endogenous influences, such as fixed assortment lengths or maximum small-end diameters, cannot be considered in the model. It enables the calculation of the maximal tree-specific wood amount but it does, of course, not predict actual decisions.

\subsection{Flexible Global Optimization with Simulated-Annealing}
\label{subsec:discussion:struct:opt}
The optimization of forest stand treatments makes it possible to determine those forest operations with the highest marginal return (Chapter \ref{chap:opt}). The developed simulation-optimization model can be used to calculate the full wood potential of an entire forest enterprise in a time period of between 10 and 20 years. It is, in contrast to the other models mentioned, a model for the support of intermediate-term decisions on a lager spatial scale. Bio-economy companies often need a continuous raw material supply to ensure a continuous production. Contractually agreed continuous wood supply quantities can have consequences for the success of upcoming bio-economy companies \citep[p. 221, 223]{elchichakli_2016}. If delivery agreements with upcoming bio-economy companies, or any other recipient, lead to opportunity costs, source providers must weight the advantages and disadvantages very carefully. The optimization model enables the calculation of those opportunity costs. The decision-maker can use the result to balance between the drawbacks of delivery contracts and their benefits in terms of planning security. They can decide whether the benefits in intermediate-term planning justify the opportunity costs. Furthermore, wood demander can use the results to examine adequate wood prices. The calculations can build an objective basis for negotiation of intermediate-term delivery contracts between forest enterprises and bio-economy companies. The software can, furthermore, be beneficial for any other analysis where minimum harvesting amounts are relevant. As forest owners usually have continuous fixed costs, they are interested in continuous yields to keep their enterprises solvent \citep[p. 74]{mohring_2010b}. Forest owners can hence use the software to find a trade-off between the advantages and drawbacks of continuous yields. A further example where the introduced simulation-optimization approach might be useful is the evaluation of future investments. If investments led to continuous costs, which may result in the need for higher yields to tackle those costs, the model could be used to calculate the opportunity costs of the implied higher yields. The model could be used to calculate the opportunity costs of future investments for forest owners, if those investments implied minimum yields within a finite time period between 5 and 20 years. The model could therefore support the trade-off process between the advantages and drawbacks of future investments.

It was shown that optimization of the harvesting schedule can be advantageous from perspective of forestry and from perspective of wood processing companies (Subsection \ref{subsec:opt:appendix:results}). As the implemented loss function in the simulation-optimization software returns the rated wood volume (Subsection \ref{subsec:intro:struct:opt}), the actual optimization process optimizes the harvesting operation planning in terms of revenues and processing costs. Decision makers who follow the optimized stand development schedules will optimize their monetary return within the simulation period, though the actual revenue is not explicitly shown because the return of the loss function is, for technical reasons, normalized. The actual result of the loss function allows therefore no direct interpretable since it is a one to one transformation of the simulated and rated wood volume.

The implemented sustainability restrictions focus only on the sustainability of the stand volumes. Many other definitions are possible \citep[p. 102]{spellmann_2010}. To interpret the stand volumes of a forest development simulation with standard treatment settings as the sustainable stand volume (Subsection \ref{subsec:intro:struct:opt}) is a model simplification. The idea behind this implementation is that sustainable stand volumes are not static. The stand specific sustainable wood potential depends on factors like tree species and tree age (Figure \ref{fig:hzb:fig4_zuw_nutz}). Advantage of this definition is that the stand specific species and age characteristics directly influence the sustainability restrictions and hence the limits for the optimization process. Disadvantage of this implementation is that the users of the simulation-optimization software must predefine simulation settings which are assumed to lead to a sustainable stand development. The parameterization of a standard scenario is hence a crucial step in the optimization procedure. \citet[p. 90]{hansen_2014} give an overview over the possible settings. In the specific examples of this thesis (Subsections \ref{subsec:opt:examples:forest} and \ref{subsec:opt:appendix:results}), the usual treatment settings of central Germany are expected to represent such a sustainable forest development. The settings based on former cluster studies for the forest development in Germany and were specifically adopted to the typical forest developments in central Germany (Section \ref{sec:hzb:Methodik}; Subsection \ref{subsec:opt:appendix:method}).

Owing to the strategic orientation of most German forest enterprises, aim of the forest management in typical European beech stands, especially in public-owned forests, in Germany is to produce high-valued stem wood \citep{nagel_2008}. In the simulation-optimization software, stem wood volume is rated higher than wood volume of lower-valued assortments like industrial wood. The simulation-optimization model hence automatically supports stem wood production since stem wood is favorable in terms of the revenue, just as it is usually the case in practical forest planning. The simulation-optimization software, in its recent form, therefore mainly optimizes the schedule and the intensity of thinning activities rather than the harvesting intensity of target usage operations. This behavior is confirmed in the second case study (Subsection \ref{subsec:opt:appendix:results}). It was nevertheless shown in two examples that optimization of thinning activities substantially increased the harvestable wood amount. The forecasted overall harvestable wood volume remarkably increased whereas the amount of high-values wood from target usage activities was nearly unchanged. Forest enterprises can, to conclude, optimize their monetary return in the simulation period with the introduced optimization software. This optimization is mainly reasoned by the rearranging of the thinning harvesting schedule.

For the same reasons, the software is also interesting for wood processing companies who are interested in this low-valued wood. It was examined that low-valued European beech wood could be sparse in the future because the demand for wood as natural resource increases while the available wood potential decreases due to nature conservation issues (Chapter \ref{chap:hzb}). Integration of the simulation-optimization software into the intermediate-term forest planning could increase the available wood potential for upcoming bio-economy companies. Following the thinning strategies of the optimized treatment scenario leads to higher wood potentials for the wood-based bio-economy without influencing the production of high-valued stem wood from target usage.

The combined si\-mu\-la\-tion-op\-ti\-mi\-za\-tion method makes high demands on the optimization algorithm (Subsections \ref{subsec:intro:struct:opt} and \ref{subsec:opt:examples:forest}). As a robust optimization module is a prerequisite for a reliable si\-mu\-la\-tion-op\-ti\-mi\-za\-tion software, several optimization approaches were tested and evaluated. Stochastic methods with dynamic random structures performed best. Only such models were able to cope with the specific requirements of the combined si\-mu\-la\-tion-op\-ti\-mi\-za\-tion method. The best optimization model found was a composition of three optimization strategies \citep{corana_1987, kirkpatrick_1983, pronzato_1984}. All approaches were separately available as software packages. A composition of the methods was, however, unpublished. The development of a specific software package for the optimization of forest thinning activities was therefore straightforward. An intensive sensitivity analysis confirmed the applicability and the efficiency of the package as the optimization element in the si\-mu\-la\-tion-op\-ti\-mi\-za\-tion software.

\section{Outlook}
\label{sec:discussion:outlook}
The next step will be to implement the developed and evaluated models into software for use by scientists, students and practitioners. The WaldPlaner DSS provides a favorable front-end for the models, as it is already established in forest sciences and practice.

One common aim of all the introduced methods is the assessment of the full wood potential. All the models provide decision support to strengthen the economic forest function. There are, however, numerous good reasons not to fully exploit this potential. Conservation or recreational issues, which could decrease the actual harvestable wood volume, cannot be considered at present. Combinations of the models from Chapters \ref{chap:bm} and \ref{chap:beech_crowns} with other DSS could, however, enable consideration of other forest functions. Already existing forest DSS might reduce the wood potential, that is predicted by one of the introduces methods, due to conservation or recreational issues. These aspects provide further arguments for an implementation of the introduced models into the WaldPlaner as the WaldPlaner allows parametrization of treatments with reduced wood potential \citep[p. 90-93]{hansen_2014}. Growth and yield simulations with nature conservation oriented treatment settings will forecast lower yields than economic oriented treatments. Combination of the presented methods from Chapters \ref{chap:bm} and \ref{chap:beech_crowns} with the WaldPlaner hence enables forecasting of the stand specific wood potential with regard to economic and nature conservation issues.

The si\-mu\-la\-tion-op\-ti\-mi\-za\-tion software (Chapter \ref{chap:opt}) optimizes the stand development with a view to economic issues. Consideration of the other two forest functions is nevertheless possible. As it was performed by \citet{yousefpour_2009}, conservation and recreational issues can be simplified such that they can be implemented as restrictions into the si\-mu\-la\-tion-op\-ti\-mi\-za\-tion software. Restrictions, such as harvesting permissions of old deciduous trees or minimum stand volumes for specific stand ages, could be included to enhance the forecasting accuracy in forest stands with specific nature conservation regulations. Such applications would enable the calculation of opportunity costs of specific conservation and recreational issues.

The influence of wood quality on the optimal stand treatment is another relevant issue that could be examined with the si\-mu\-la\-tion-op\-ti\-mi\-za\-tion software. To achieve this, an interface with user-specific wood prices for different wood assortments must be implemented.

Cooperation between wood-processing companies themselves, as well as between the companies and forest enterprises, is one of the most important advantages of the bio-economy sector (Section \ref{sec:intro:biecon}). The assessment and interpretation of complete raw material supply chains appears to be especially interesting. The methods introduced in this thesis were developed to help strengthen decisions in specific parts of such supply chains, in order to promote the cooperation of the forest and the bio-economy sector. The statistical methods, however, cover only small, distinct parts of the supply chains. To enable meaningful cooperation, they must be somehow made applicable for decision-makers in forestry, as well as in bio-economy or in logistic companies. When implementing the models into DSS, interfaces to other software will be mandatory in order to take full advantage of their possibilities. A front-end with an interface to logistic DSS could link the optimized wood amounts with resource distribution simulations or optimizations.